Derive final annotations based on NBAtlas reference

### 2025-08-19 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
add testing flag for CI since dotplot fails, and update notebook text to match README

---
title: "Derive final annotations based on NBAtlas reference"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  scanvi_pp_threshold: 0.75    
  annotations_tsv: "../results/SCPCP000004-annotations.tsv"
---

@@

We use the following approach to determine the final annotation.

First, we prioritize inferences with `SingleR` and `scANVI/scArches`.
If these methods returned the exact same label, we assign this label.

Next, we compare labels across all three methods: if at least _two out of three_ methods agree, we assign the agreeing label.
Note that this requires mapping consensus cell types to NBAtlas labels, which is only possible for these cell types: endothelial cells, fibroblasts, and immune cell types except for the NBAtlas labels `RBCs` and `pDC` which do not have corresponding consensus labels.

In this approach, there are two ways that cell type annotations can agree, and we treat them as follows:

* Cell type annotations can have directly corresponding labels. For example:
  * If the consensus cell type is `endothelial cell` and at least one of the NBAtlas-derived annotations is `Endothelial`, we assign the label `Endothelial`.
* Cell type annotations can differ but be in the same "family" of cell types. For example:
  * If the consensus cell type is `mature T cell` and at least one of the NBAtlas-derived annotations is `CD4+ T cell`, we assign the broader label `T cell`.
  * If one NBAtlas-derived annotation is `Patrolling monocyte` and the other is `Classical monocyte`, we assign the broader label `Monocyte`.

We use CL ontology ids to determine agreement throughout. 
For more information about how these ontology ids were determined for NBAtlas labels, see `../references/README.md`.

For the final exported TSV file, we further designate any cells labeled as `Neuroendocrine` as `tumor`, following how this label is interpreted in NBAtlas.

@@

```{r, fig.width = 32, fig.height = 16}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  celltype_df = input_dotplot_df,
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = celltype_pal, 
  min_cells = 10 # only show cell types with at least 10 cells
) +
  theme(legend.position = "bottom")
```

>>>>

---
title: "Derive final annotations based on NBAtlas reference"
author: "Stephanie J. Spielman"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 3
    code_folding: hide
params:
  scanvi_pp_threshold: 0.75    
  annotations_tsv: "results/SCPCP000004-annotations.tsv"
  testing: FALSE
---

@@

We use the following approach to determine the final annotation.
Throughout, we use CL ontology ids to perform comparisons. 
For more information about how these ontology ids were determined for NBAtlas labels, see `../references/README.md`.

1. If `SingleR` and `scANVI/scArches` labels exactly agree, we assign that label
2. If broad label families (e.g., `T cell` would be the broad family for `CD4+ T cell`) for `SingleR` and `scANVI/scArches` labels agree, we assign the broad family label.
  * See [`references/nbatlas-label-map.tsv`](references/nbatlas-label-map.tsv) and [`references/README.md`](references/README.md) for how labels were mapped to families.
3. If `SingleR` and `scANVI/scArches` labels disagree, we then compare to the consensus cell type; if at least one inference agrees with the consensus cell type, we assign the inferred label.
  * We perform this first at the label level and then at the broad family level.
4. Finally, following the conclusions of Bonine et al. (2024), we designate any cell labeled as `Neuroendocrine` as a tumor cell.

@@

```{r, fig.width = 32, fig.height = 16, eval = !params$testing}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  celltype_df = input_dotplot_df,
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = celltype_pal, 
  min_cells = 10 # only show cell types with at least 10 cells
) +
  theme(legend.position = "bottom")
```
